---
title: "16: Writing Quantitative Social Science"
subtitle: "Statistical Research Methods"
author: "<large>J. Seawright</large>"
institute: "<small>Northwestern Political Science</small>" 
date: "May 21, 2025"
output: 
  xaringan::moon_reader:
    css: xaringan-themer.css
  
---
class: center, middle

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
}
```

```{css, echo=FALSE}
.scroll-100 {
  max-height: 100px;
  overflow-y: auto;
  background-color: inherit;
}
```

```{r, load_refs, include=FALSE, cache=FALSE}
# Initializes the bibliography
library(RefManageR)

library(ggplot2)
library(dplyr)
library(readr)
library(nlme)
library(jtools)
library(hrbrthemes)
library(mice)
library(knitr)
options(warn=-1)

BibOptions(check.entries = FALSE,
           bib.style = "authoryear", # Bibliography style
           max.names = 3, # Max author names displayed in bibliography
           sorting = "nyt", #Name, year, title sorting
           cite.style = "authoryear", # citation style
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE)
#myBib <- ReadBib("assets/myBib.bib", check = FALSE)
# Note: don't forget to clear the knitr cache to account for changes in the
# bibliography.
```
```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer,MnSymbol)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  text_font_size = "1.3rem"
)
```


---
### Outline

-   Introduce the research question and briefly explain why it is important.

-   Explain why existing research is incomplete.

-   Describe your data and tell where you found it.

-   Describe your statistical analysis and explain the assumptions you're making.

-   Present your results.

-   Explain what we have learned, and point toward next steps to learn more.

---
### Make sure to...

-   Report unanticipated events that occurred during your data collection. 

-   Explain how the actual analysis differs from the planned analysis. 

-   Explain your handling of missing data. 

-   Be clear about any limitations to your analysis.

---
### Presenting Numbers

Use tables to provide exact values; use figures to convey global effects. 

Keep figures small in size; include graphic representations of confidence intervals whenever possible.

---
### Interpreting Tables and Figures

-   Always tell the reader what to look for in tables and figures. 

-   Discuss the specific numbers that matter and their meaning. 

-   A reader who skips the tables and figures should still be told the statistical results.
