---
title: "Choosing Control Variables: DAGs & the Back-Door Criterion"
subtitle: "PS 312"
author: "Jaye Seawright"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(broom)
library(ggplot2)
library(ggdag)
library(dagitty)
theme_set(theme_dag())
```

## Today's Roadmap

1. **Hook & Activation:** Quick review of back‑door paths (poll)
2. **"Bad Control" Diagnostic:** Three scenarios to sharpen your instincts
3. **Variable Dump Challenge:** Choose controls from a real dataset
4. **Case Study Deconstruction:** Analyze a published regression table
5. **Core Graded Activity:** Group DAG creation and justification

**Goal:** Move from "I can draw a DAG" to "I know exactly which variables belong in my regression—and which don't."

---

class: inverse, center, middle

# 1. Hook & Activation  
### Quick Review of the Back‑Door Criterion

---

## Recall from your Preparation

A set of variables \( Z \) satisfies the **back‑door criterion** for \((X, Y)\) if:

1. No node in \( Z \) is a *descendant* of \( X \).
2. \( Z \) blocks **every** path between \( X \) and \( Y \) that starts with an arrow pointing *into* \( X \).

---

> **Simple check:** Does adjusting for \( Z \) close all sneaky back‑door paths without opening any new ones?

---

## Let's Discuss

**DAG:**  
`Parental Income` → `Education` → `Earnings`  
`Parental Income` → `Earnings`

**Question:** To estimate the *total* effect of `Education` on `Earnings`, should we control for `Parental Income`?

- A) Yes, it's a confounder.
- B) No, it's a mediator.
- C) Yes, but only if we also control for something else.
- D) It doesn't matter.

---

class: inverse, center, middle

# 2. "Bad Control" Diagnostic  
### Three Scenarios to Test Your Instincts

---

## Scenario 1:

```{r dag-step1, echo=FALSE, fig.width=5, fig.height=3}
dag_mediator <- dagify(
  Turnout ~ Interest,
  Interest ~ Contact,
  exposure = "Contact",
  outcome = "Turnout",
  labels = c("Contact" = "Campaign\nContact",
             "Interest" = "Political\nInterest",
             "Turnout" = "Voter\nTurnout"),
  coords = list(
    x = c(Contact = 0.3, Interest = 1.5, Turnout = 2.7),
    y = c(Contact = 0.3, Interest = 0.7, Turnout = 0.3)
  )
)

ggdag_status(dag_mediator, text = FALSE) +
  geom_dag_label(aes(label = label), parse = FALSE,
                  nudge_x = c(0, 0, 0),
                  nudge_y = c(-0.05, 0.05, -0.05),
                  size = 4) +
  labs(title = "DAG")
```

**Question:** You want the *total* effect of Campaign Contact on Turnout. Should you control for **Political Interest**?

- [ ] Yes  
- [ ] No

**Why?**

---

## Scenario 2: 

```{r dag-step2, echo=FALSE, fig.width=5, fig.height=3}
dag_collider <- dagify(
  Star ~ Talent,
  Star ~ Charisma,
  exposure = "Charisma",
  outcome = "Talent",
  labels = c("Star" = "Star\nStatus",
             "Talent" = "Talent",
             "Charisma" = "Charisma"),
  coords = list(
    x = c(Talent = 0.3, Star = 1.5, Charisma = 2.7),
    y = c(Talent = 0.3, Star = 0.7, Charisma = 0.3)
  )
)

ggdag_status(dag_collider, text = FALSE) +
  geom_dag_label(aes(label = label), parse = FALSE,
                  nudge_x = c(0, 0, 0),
                  nudge_y = c(-0.05, 0.05, -0.05),
                  size = 4) +
  labs(title = "DAG")
```


**Question:** You are studying whether *Charisma* affects *Talent*. Should you control for **Star Status**?

- [ ] Yes  
- [ ] No

**Why?**

---

## Scenario 3: 

```{r dag-step3, echo=FALSE, fig.width=5, fig.height=3}
dag_colliders <- dagify(
  Corruption ~ Pastscandals + Wealth,
  Multiculturalism ~ Highereducation + Wealth,
  exposure = "Corruption",
  outcome = "Multiculturalism",
  labels = c("Pastscandals" = "Prior\nScandals",
             "Corruption" = "Corrupt\nReputation",
             "Wealth" = "Wealth",
             "Highereducation" = "Higher\nEducation",
             "Multiculturalism" = "Multiculturalism"),
  coords = list(
    x = c(Corruption = 0.5, 
          Pastscandals = 1.5, 
          Wealth = 3.0, 
          Highereducation = 4.5, 
          Multiculturalism = 5.5),
    y = c(Corruption = 0.8,
          Pastscandals = 1.3,
          Wealth = 0.8,
          Highereducation = 1.3,
          Multiculturalism = 0.8)
  )
)
ggdag_status(dag_colliders, text = FALSE) +
  geom_dag_label(aes(label = label), parse = FALSE,
                  nudge_x = c(0, 0, 0, 0, 0),
                  nudge_y = c(-0.05, 0.05, -0.05, 0.05, -0.05),
                  size = 4) +
  labs(title = "DAG")
```


**Question:** Is **Wealth** a confounder? Should you include it?

- [ ] Yes  
- [ ] No

**Why?**

---

class: inverse, center, middle

# 3. Variable Dump Challenge  

---

## The Research Question

> Does attending religious services increase an individual's annual charitable giving (in dollars)?

**Data source:** General Social Survey (GSS)

**Main variables:**
- **Treatment:** `Religious Attendance` (0 = never, 8 = weekly+)
- **Outcome:** `Charitable Giving` (total dollars donated to non‑religious charities)

---

## Available Variables in the Dataset

- `age` (years)
- `income` (annual household income)
- `educ` (years of education)
- `parent_relig` (how often parents attended services when R was child)
- `altruism` (self‑reported agreement with "I often try to help others")
- `pol_ideology` (1=extremely liberal to 7=extremely conservative)
- `married` (1=married, 0=otherwise)
- `childs` (number of children)
- `south` (1=lives in South, 0=otherwise)

---

## Group Tasks (10 minutes)

1. **Draw a minimal DAG** on scratch paper. Include at least `Religious Attendance`, `Charitable Giving`, and two other variables from the list that you suspect are confounders.

2. **Select TWO variables** from the list that you would **definitely include as controls** in a regression.

3. **Select ONE variable** from the list that you are **actively deciding NOT to control for**, even though it's available.

**We will discuss your decisions together as a group at the end of 10 minutes.**  

---

## Discussion Debrief

Let's hear from a few groups:

- Which two variables did you pick as controls? Why are they confounders?
- Which variable did you **not** control for? Why?

**Watch out for:**  
- `altruism` → likely a **mediator** (attendance → altruism → giving)  
- `parent_relig` → strong **confounder** (upbringing affects both attendance and giving norms)  
- `income` → **confounder** (ability to give and possibly attendance patterns)

---

class: inverse, center, middle

# 4. Case Study Deconstruction  
### DAG Thinking in the Wild

---

## Excerpt from a Published Study

> **Research Question:** Does receiving mail that publicizes one's voting record (treatment) increase voter turnout (outcome)?

**Study:** Gerber, Green, and Larimer (2008) – *Social Pressure and Voter Turnout*

**Design:** Large‑scale field experiment with random assignment to treatment conditions.

---

## Regression Table (Simplified)

| Variable | Model 1 (No Controls) | Model 2 (With Controls) |
| :--- | :--- | :--- |
| Treatment (Social Pressure Mail) | 0.081*** (0.008) | 0.076*** (0.008) |
| Age | | 0.004*** (0.000) |
| Party Registration (Democrat) | | 0.012 (0.010) |
| **Voted in Previous Election** | | 0.342*** (0.010) |
| Household Size | | -0.002 (0.003) |
| *Constant* | 0.302*** | 0.085*** |
| Observations | 80,000 | 80,000 |

*Standard errors in parentheses. ***p < 0.01.*

---

## Discussion Questions

1. **Coefficient change:** The treatment effect goes from 0.081 to 0.076. Is this a **large** change? What does it tell us about the relationship between treatment and the added controls?

2. **Past turnout as a control:** Is `Voted in Previous Election` a **confounder** in this experiment? Why or why not?

3. **Why include it then?** If treatment was randomly assigned, why might the authors still include `Voted in Previous Election`? What benefit does it provide?

4. **A thought experiment:** Suppose the treatment was **not** randomized but instead a "Get Out the Vote" phone call that campaigns target at likely voters. How would your answer to Q2 change?

---

## Key Takeaways from the Case Study

- **In randomized experiments,** covariates are *not* needed for unbiased estimation.  
- **But** including strong predictors of the outcome (like past turnout) **reduces standard errors** and improves precision.
- **In observational studies,** variables like past turnout are often **confounders** that *must* be controlled to avoid bias.

> The back‑door criterion tells you **when** to control.  
> Precision considerations tell you **why** you might still control even when unconfounded.

---

class: inverse, center, middle

# 5. Core Graded Activity  
### Group DAG Creation & Justification

---

## Your Task (20 minutes)

In your assigned group, create a DAG for a research question of your choice (or use the one provided below).

**Requirements:**

1. **Draw the DAG** (on paper, whiteboard, or using DAGitty). Take a photo/screenshot.
2. **Identify at least three variables** that are **confounders** for the main relationship. These should be variables you would include as controls.
3. **Explain why one of these variables meets the definition of a confounder.** (1–2 sentences: "Variable Z is a confounder because it is a common cause of both X and Y...")
4. **Identify one variable that you explicitly chose NOT to include** as a control. Explain why it does *not* meet the definition (e.g., it's a mediator, collider, or post‑treatment).

---

## Submission

**Each student** must submit to the TA via email **by the end of today**:

- The DAG image (photo/screenshot)
- The written justifications (confounder explanation + omitted variable explanation)

---

class: inverse, center, middle

# Wrap‑Up

---

## What's the Single Most Important Rule?

> **"Close all back‑door paths. Do not control for mediators, colliders, or post‑treatment variables."**

If you remember only one thing from today, make it this.
