---
title: "Class Activity: DAG Construction"
subtitle: "PS 312"
author: "Jaye Seawright"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dagitty)
library(ggdag)
library(ggplot2)
```

## Today's Activity

- **Goal:** Translate your team's research question into a **Directed Acyclic Graph (DAG)**.
- **Deliverable:** Each group member sends the DAG to their TA with a short paragraph explaining the nodes and edges.
- **Why?** DAGs make causal assumptions transparent and guide covariate selection.

I'll walk through an example from my own work on the January 6 insurrection, building the DAG step by step.

---

## Step 1: The Core Relationship

**Research Question:** Does social media use increase the likelihood of engaging in violence during the January 6 Capitol attack?

We start with the simplest possible DAG: just the exposure and the outcome.

```{r dag-step1, echo=FALSE, fig.width=5, fig.height=3}
dag1 <- dagitty('dag {
  SocialMedia [pos="0,0"]
  Violence [pos="2,0"]
  SocialMedia -> Violence
}')

ggdag(dag1, layout = "nicely", text_col = "blue") +
  theme_dag() +
  labs(title = "Step 1: Core Relationship")
```

- **SocialMedia:** Use of platforms like Twitter, Facebook, Parler, YouTube to live broadcast the experience of January 6th.
- **Violence:** Physical aggression during the Capitol breach (assault, property destruction).

This is our starting hypothesis. But is this relationship causal, or could something else explain it?

---

## Step 2: Adding Confounders

To estimate the causal effect of SocialMedia on Violence, we need to consider **confounders**—variables that affect both the exposure and the outcome.

```{r dag-step2, echo=FALSE, fig.width=8, fig.height=5}
dag2 <- dagitty('dag {
  SocialMedia [pos="0,0"]
  Violence [pos="2,0"]
  SocialMedia -> Violence
  
  PoliticalEngagement [pos="0,2"]
  InternetUse [pos="0,1"]
  Personality [pos="-1,1"]
  Age [pos="-1,0"]
  Education [pos="-1,-1"]
  Region [pos="0,-1"]
  
  PoliticalEngagement -> SocialMedia
  PoliticalEngagement -> Violence
  InternetUse -> SocialMedia
  InternetUse -> Violence
  Personality -> SocialMedia
  Personality -> Violence
  Age -> SocialMedia
  Age -> Violence
  Education -> SocialMedia
  Education -> Violence
  Region -> SocialMedia
  Region -> Violence
}')

ggdag(dag2, layout = "nicely", text_col = "blue") +
  theme_dag() +
  labs(title = "Step 2: Add Confounders")
```

---

**Confounders (each affects both SocialMedia and Violence):**
- **PoliticalEngagement:** Activists use social media more and attend more protests (some turn violent).
- **InternetUse:** Tech‑savvy individuals use social media heavily and were over‑represented in D.C.
- **Personality:** Traits like sensation‑seeking or aggression.
- **Age:** Younger people use social media more and may be more impulsive.
- **Education:** Correlated with digital literacy and political participation.
- **Region:** Areas with strong pro‑Trump sentiment had higher social media engagement and rally attendance.

**Implication:** To estimate SocialMedia → Violence without bias, we must control for these confounders.

---

## Step 3: Downstream Consequences

Some participants may have been motivated by a desire for **media visibility**. Does engaging in violence increase post‑attack fame?

```{r dag-step3, echo=FALSE, fig.width=9, fig.height=6}
dag3 <- dagitty('dag {
  SocialMedia [pos="0,0"]
  Violence [pos="2,0"]
  SocialMedia -> Violence

  PoliticalEngagement [pos="0,2"]
  InternetUse [pos="0,1"]
  Personality [pos="-1,1"]
  Age [pos="-1,0"]
  Education [pos="-1,-1"]
  Region [pos="0,-1"]
  
  PoliticalEngagement -> SocialMedia
  PoliticalEngagement -> Violence
  InternetUse -> SocialMedia
  InternetUse -> Violence
  Personality -> SocialMedia
  Personality -> Violence
  Age -> SocialMedia
  Age -> Violence
  Education -> SocialMedia
  Education -> Violence
  Region -> SocialMedia
  Region -> Violence

  MediaVisibility [pos="4,0"]
  Violence -> MediaVisibility
  PriorFame [pos="3,1"]
  PriorFame -> MediaVisibility
  PriorFame -> Violence [pos="3,0.5"]
}')

ggdag(dag3, layout = "nicely", text_col = "blue") +
  theme_dag() +
  labs(title = "Step 3: Add Downstream Outcome (Media Visibility)")
```

---

**New Nodes:**
- **MediaVisibility:** Post‑attack media coverage (news articles, interviews, social media fame).
- **PriorFame:** Pre‑existing public profile (e.g., local politician, internet personality).

**New Edges:**
- **Violence → MediaVisibility:** Violent acts attract coverage—this could be a motive.
- **PriorFame → MediaVisibility:** Famous people get more coverage regardless.
- **PriorFame → Violence:** (Dashed) Possible that prior fame increases propensity to act out.

**Analytical Note:** If we want to estimate Violence → MediaVisibility, we must control for PriorFame (a confounder of that relationship).

---

## How This DAG Informs Analysis

- For **SocialMedia → Violence:** Control for PoliticalEngagement, InternetUse, Personality, Age, Education, Region.
- **Do NOT** control for MediaVisibility (it's downstream).
- For **Violence → MediaVisibility:** Control for PriorFame; consider instrument or natural experiment for Violence.

---

## Your Turn

1. Sketch a DAG for your team's research question, starting simple and building up.
2. Label each node and explain it. Your entire explanation need not exceed a paragraph.
3. Identify:
   - The main variables that represent your causal question
   - Background variables that help us understand the situation and meaningful differences among cases
   - Variables that help tell the causal story of how effects might work
   - Downstream variables that help interpret the causal question
4. Send your DAG (image or R code) and paragraph to your TA.

**Resources:**
- [Class DAG Help Page](https://jnseawright.github.io/PS312/InClass/Causation_Causal_Inference.html)
- `dagitty` and `ggdag` R packages

---

```{r, echo=TRUE, eval=FALSE}
# Example R code for students:
library(dagitty)
my_dag <- dagitty('dag {
  X -> Y
  Z -> X
  Z -> Y
}')
plot(my_dag)
```
