---
title: "Instrumental Variables"
subtitle: "PS 312"
author: "Jaye Seawright"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(AER)         # ivreg() with diagnostics
library(knitr)
```

## Today's Roadmap

1. **Hook & Activation:** The logic of instrumental variables  
2. **Concept Introduction:** The IV model, relevance, and exclusion  
3. **Instrument on Trial** – Does rainfall pass the test?  
4. **Diagnostics & Assumptions:** Weak instruments and overidentification  
5. **Instruments Brainstorm:** Inventing instruments that might work in political science  
6. **Core Graded Activity:** Write your paragraph for the TA  
7. **Wrap‑Up:** Cheat sheet for IV designs  

**Goal:** Move from "I've heard of IV" to "I can design, run, and critically evaluate an instrumental‑variables analysis in R."

---

class: inverse, center, middle

# 1. Hook & Activation  
### The Logic of Instrumental Variables

---

## Scenario: Returns to Education

You want to know the causal effect of an extra year of schooling on wages. You run a regression:

`$$\text{Wage}_i = \beta_0 + \beta_1 \text{Education}_i + u_i$$`

- Why might `\(\beta_1\)` be biased? (Hint: Ability is unobserved and correlated with both education and wages.)

If we could find a variable that **affects education** but **has no direct effect on wages** (other than through education), we could use it as an **instrument**. A famous example is **quarter of birth**: people born in different quarters have slightly different compulsory schooling lengths, but birth quarter should not directly affect wages decades later.

---

## The IV Intuition

1. **Instrumental variable (Z):** A variable that is correlated with the endogenous regressor (`X`) but uncorrelated with the error term (`u`).  
2. **First stage:** Regress `X` on `Z` (and any exogenous covariates) to get the predicted values `\(\hat{X}\)`.  
3. **Second stage:** Regress `Y` on `\(\hat{X}\)` (and the covariates). The coefficient on `\(\hat{X}\)` is the **IV estimate**.

The IV estimator isolates variation in `X` that is *as good as random* (driven by `Z`), purging the bias from unobserved confounders.

---

class: inverse, center, middle

# 2. Concept Introduction  
### The IV Model, Relevance, and Exclusion

---

## The Two IV Assumptions

1. **Relevance:** The instrument must be strongly correlated with the endogenous regressor.  
   - *Check:* First‑stage F‑statistic > 10 (rule of thumb).
2. **Exclusion restriction:** The instrument must affect the outcome **only through** the endogenous regressor.  
   - *Cannot be formally tested*—must be defended theoretically.

For our activity, we will use **election‑day rainfall and snowfall** as instruments for **voter turnout**. Do these instruments satisfy the assumptions?

---

## The 2SLS Estimator in R

We will use the `ivreg()` function from the `AER` package. The formula syntax is:

`y ~ x1 + x2 | z1 + z2 + x2`

- `y` is the outcome.
- `x1` is the **endogenous** regressor (turnout).
- `x2` are **exogenous** covariates (included in both stages).
- `z1` and `z2` are the **instruments** (rain, snow).

`ivreg()` automatically performs two‑stage least squares and can report diagnostics with `summary(..., diagnostics = TRUE)`.

---

class: inverse, center, middle

# 3. Instrument on Trial  
### Does Rainfall Pass the Test?

---

## The Setup

You've seen the code that loads the data and runs the IV. Now we put the instruments—`Rain` and `Snow`—**on trial**.

| **Role** | **Your Task** |
| :------- | :------------ |
| **Defense** | Argue that `Rain` and `Snow` are **valid instruments** for `Turnout`. Use the regression output to support your case. |
| **Prosecution** | Argue that `Rain` and `Snow` are **invalid instruments**. Identify violations of relevance or the exclusion restriction. |

**The evidence:** The R output from the analysis (OLS, IV, first‑stage, Sargan test).

---

**Procedure:**
1. (3 min) We need volunteers to act as the prosecution and the defense. Each side will have a few minutes to prepare and to meet with the instructor.
2. (5 min) Defense presents first; Prosecution rebuts.
3. (2 min) Class votes: **Valid** or **Invalid**?

---

## Loading the Data

The dataset `rainfallections.csv` contains county‑level election returns, battleground status, and weather for U.S. presidential elections from 1948 to 2000.

```{r}
iv_data <- read_csv("data/rainfallelections.csv")

# Keep relevant variables and remove missing
iv_clean <- iv_data %>%
  select(Year, State, County, Turnout, battleground, Rain, Snow) %>%
  filter(!is.na(Turnout), !is.na(battleground), !is.na(Rain), !is.na(Snow))

glimpse(iv_clean)
```

---

## Exhibit A: The OLS Regression

```{r}
ols_model <- lm(battleground ~ Turnout, data = iv_clean)
summary(ols_model)
```

---

## Exhibit B: The IV Regression (2SLS)

```{r}
iv_model <- ivreg(battleground ~ Turnout | Rain + Snow,
                  data = iv_clean)
summary(iv_model, diagnostics = TRUE)
```
---

**Defense highlights:**
- First‑stage F‑statistic > 10 → instruments are **relevant**.
- Wu‑Hausman test significant → turnout is indeed endogenous; IV is justified.

**Prosecution highlights:**
- Sargan test p‑value is small → **overidentification rejected**; at least one instrument may be invalid.
- The coefficient changed quite a bit from OLS, which could be a sign of weak‑instrument bias even if F > 10.

---

## Exhibit C: The First Stage (Relevance)

```{r}
first_stage <- lm(Turnout ~ Rain + Snow, data = iv_clean)
summary(first_stage)
```

---

## Exhibit D: The Exclusion Restriction (Conceptual)

*No table for this one—it's about theory.*

**Defense needs to argue that...:**
> Rainfall on Election Day affects battleground status **only** by reducing turnout.

**Prosecution needs to argue that...:**
> Weather could affect long‑term political competitiveness in some way other than by reducing turnout. (Ideally, provide one or more examples of such an effect.) 

---

## Your Verdict

After hearing both sides, **raise your hand**:

- ✋ **Valid:** The instruments are credible.
- ✋ **Invalid:** The instruments fail the relevance or exclusion requirement.

*(Instructor tallies and discusses the majority view.)*

---

## Why This Matters for Your Project

In your own analysis (using `GOPVoteShare` as the outcome), you will face the **exact same debate**. You must decide whether to trust the IV results. The tools you just used—first‑stage F, Sargan test, theoretical reasoning—are what you'll apply to defend your own conclusions.

**Remember:** No instrument is perfect. The goal is to be **transparent** about the threats and to show that your results are robust to reasonable challenges.

---

class: inverse, center, middle

# 4. Diagnostics & Assumptions  
### Weak Instruments and Overidentification

---

## Weak Instrument Test

A weak instrument leads to biased IV estimates and inflated standard errors. The rule of thumb: **first‑stage F‑statistic > 10**.

We can examine the first stage directly (as in Exhibit C) or extract the F‑statistic from the `ivreg` diagnostics.

```{r}
# Extract first‑stage F‑statistic from ivreg object
summary(iv_model, diagnostics = TRUE)$diagnostics["Weak instruments", "statistic"]
```

If this value is less than 10, the instruments are **weak** and the IV estimates should be interpreted with caution.

---

## Overidentification Test (Sargan)

When we have **more instruments than endogenous regressors**, we can test whether the instruments are uncorrelated with the second‑stage error term.

```{r}
summary(iv_model, diagnostics = TRUE)$diagnostics["Sargan", "p-value"]
```

- **Null hypothesis:** The instruments are valid (exclusion restriction holds).
- A small p‑value (< 0.05) suggests that at least one instrument may be invalid.

---

## Robustness Checks

Do the results change if we use only one instrument (`Rain`)? What if we add controls like year fixed effects or state fixed effects?

```{r}
# IV with only rain
iv_rain <- ivreg(battleground ~ Turnout | Rain, data = iv_clean)
summary(iv_rain, diagnostics = TRUE)

# With state fixed effects
iv_fe <- ivreg(battleground ~ Turnout + factor(State) | Rain + Snow + factor(State),
               data = iv_clean)
summary(iv_fe, diagnostics = TRUE)
```

---

class: inverse, center, middle

# 5. Instruments Brainstorm  
### Inventing Instruments That Might Work in Political Science

---

## The Challenge

Finding a valid instrument is **hard**. The best instruments come from **institutional quirks, natural experiments, or random assignment** that is plausibly exogenous.

In small groups, take **5 minutes** to brainstorm an instrument for one of the following research questions:

| **Research Question** | **Endogenous Regressor** | **Your Task** |
| :-------------------- | :----------------------- | :------------ |
| Does attending a political protest increase future civic engagement? | Protest attendance | Propose an instrument for protest attendance. |
| Does campaign spending increase a candidate's vote share? | Campaign spending | Propose an instrument for campaign spending. |
| Does having a female mayor increase spending on public education? | Mayor's gender | Propose an instrument for electing a female mayor. |
| Does watching partisan news make people more polarized? | Partisan news consumption | Propose an instrument for watching Fox News or MSNBC. |

---

## Some Classic Examples (For Inspiration)

| **Instrument** | **Endogenous Regressor** | **Why It Works** |
| :------------- | :----------------------- | :--------------- |
| **Vietnam War draft lottery number** | Military service | Randomly assigned; affects earnings only through service. |
| **Distance to the nearest community college** | College attendance | Affects attendance but (arguably) not wages directly. |
| **Rainfall on Election Day** | Voter turnout | Affects turnout but (arguably) not vote share directly. |
| **Quarter of birth** | Years of schooling | Affects compulsory schooling length; uncorrelated with ability. |
| **Judge's ideology (randomly assigned)** | Sentencing severity | Random assignment of cases to judges; affects outcomes only through sentencing. |

---

## Share Out

Each group will share their **best instrument idea** and defend why it satisfies:

1. **Relevance** – Strongly correlated with the endogenous regressor.  
2. **Exclusion** – Affects the outcome *only through* that regressor.

The class will vote on the **most creative but credible instrument**.

---

class: inverse, center, middle

# 6. Core Graded Activity  
### Write Your Paragraph for the TA

---

## Instructions

**By the end of class today, email your TA a short paragraph that includes:**

1. Your **research question** (one sentence).  
2. A brief description of the **IV design** (endogenous regressor, instruments, outcome).  
3. The **key result** (the IV coefficient from `ivreg()`) and how it differs from the OLS estimate.  
4. An assessment of the **instrument strength** (first‑stage F‑statistic) and what it implies.  
5. **One critical reflection** on the validity of rainfall/snowfall as instruments (exclusion restriction).

---

## Example Paragraph (for a different question)

> *Our group asks: Does education increase wages? We use quarter of birth as an instrument for years of schooling. The OLS estimate suggests a 10% return per year of education, but the IV estimate is only 7% (p < 0.01). The first‑stage F‑statistic is 12.3, above the rule‑of‑thumb threshold of 10, suggesting the instrument is adequately strong. The exclusion restriction requires that birth quarter affects wages only through schooling; while this is plausible, critics note that birth quarter may correlate with season‑of‑birth effects on health. Overall, the IV results support a causal effect of education, albeit smaller than OLS suggests.*

---

## Reminders

- One submission per student.

---

class: inverse, center, middle

# 7. Wrap‑Up  

> **Single most important rule for IV:** Always report the first‑stage F‑statistic. A weak instrument can be worse than no instrument at all.
