---
title: "Political Science 312 Homework 5"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: tango
  pdf_document: default
urlcolor: blue
---

```{r include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,                 # don't show code
  warning = FALSE,              # don't show warnings
  message = FALSE,              # don't show messages (less serious warnings)
  cache = FALSE,                # set to TRUE to save results from last compilation
  fig.align = "center",         # center figures
  attr.source = '.numberLines'  # add line numbers to code
)

library(tidyverse)       # load libraries you always use here

set.seed(27)             # make random results reproducible
```

# Due Date: May 3, 2024

### Make a DAG

Put together a graph, using daggity like the example below, to show the causal structure you think is going on behind your research question. Make sure to include your main explanatory variable, your outcome variable, any variables on the causal pathway between them, and any other variables that you think causally affect either or both of them.

Explain your reasoning behind the variables you include and the links you specify, but also the the variables and links you leave out.

```{r, echo = TRUE, out.width="100%", fig.retina = 1}
library(dagitty)
dag_cand3 <- dagitty('dag {
                          Difficulty_Conceiving -> PostnatalCare
                          Genetics -> Birth_Defects
                          Genetics -> Difficulty_Conceiving
                          PostnatalCare -> Birth_Defects
                          PostnatalCare -> Vitamins
                          SES -> PostnatalCare
                          SES -> Vitamins
                          Vitamins -> Birth_Defects
                          }')
plot(dag_cand3)
```