---
title: "PS 312: Final Presentation Evaluation Rubric"
author: "Instructor & TAs"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: false
  html_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)
library(kableExtra)
```

## Scoring Guidelines

Each criterion is scored on a **1–5 scale**. Use the descriptors below as a guide.

| Score | Description |
|:-----:|:------------|
| **5** | Excellent – exceeds expectations; clear, compelling, and rigorous. |
| **4** | Very Good – meets expectations with minor room for improvement. |
| **3** | Satisfactory – meets basic expectations; some gaps in clarity or rigor. |
| **2** | Needs Improvement – significant gaps; unclear or underdeveloped. |
| **1** | Poor – missing or fundamentally flawed. |

---

## Evaluation Rubric

```{r rubric-table, echo=FALSE}
rubric <- data.frame(
  Criterion = c(
    "1. Research Design & Causal Identification",
    "2. Assumption Evidence & Diagnostics",
    "3. Professionalism & Clarity",
    "4. Creativity & Originality",
    "5. Use & Quality of Graphs/Tables",
    "6. Handling of Q&A and Discussant Feedback"
  ),
  Weight = c("20%", "20%", "15%", "15%", "15%", "15%"),
  Excellent_5 = c(
    "Clearly identifies a causal design taught in class; articulates identification strategy and why it is credible.",
    "Presents explicit diagnostic evidence (e.g., parallel trends plot, first-stage F-stat). Discusses limitations honestly.",
    "Slides polished and well-organized; speaker maintains eye contact, speaks clearly, stays within time.",
    "Question is novel or applies classic design to fresh context; presentation uses best practices to engage the audience.",
    "Figures self-contained, properly labeled, and effectively communicate main result. Tables highlight key coefficients.",
    "Responds to questions with precision and professionalism; acknowledges limitations; engages constructively with discussant."
  ),
  Satisfactory_3 = c(
    "Mentions a design but does not fully explain identifying variation or how it addresses confounding.",
    "Mentions assumptions but provides weak or no diagnostic evidence.",
    "Slides adequate but minor issues (e.g., too much text). Speaker reads slides occasionally or rushes.",
    "Question and design are competent but generic ; presentation shows little effort to engage.",
    "Figures present but have minor labeling issues or are hard to interpret.",
    "Answers questions but may be vague or defensive; partially addresses discussant feedback."
  ),
  Needs_Improvement_1 = c(
    "No clear design; relies on cross-sectional correlations without addressing endogeneity.",
    "Ignores assumptions entirely; presents results as self-evident.",
    "Slides messy, unreadable, or disorganized. Speaker unprepared or significantly over time.",
    "Question and/or design are superficial; presentation puts obstacles in audience's path.",
    "Figures missing, misleading, or unreadable; raw output pasted onto slides.",
    "Cannot answer questions; dismisses feedback; seems unfamiliar with project."
  )
)

kable(rubric, format = "latex", booktabs = TRUE, longtable = FALSE, 
      caption = "Detailed Rubric Descriptors") %>%
  kable_styling(latex_options = c("striped", "hold_position"), font_size = 9) %>%
  column_spec(1, width = "2.8cm") %>%
  column_spec(2, width = "1.2cm") %>%
  column_spec(3:5, width = "3.5cm")
```

---

## Score Calculation

```{r score-formula, echo=FALSE}
score_calc <- "
**Final Score (out of 5)** = 
(C1 × 0.20) + (C2 × 0.20) + (C3 × 0.15) + (C4 × 0.15) + (C5 × 0.15) + (C6 × 0.15)
"
cat(score_calc)
```

Round to one decimal point before translating.

**Grade Translation (Example):**
- 4.5 – 5.0 = A
- 4.0 – 4.4 = A–
- 3.5 – 3.9 = B+
- 3.0 – 3.4 = B
- Below 3.0 = C or lower

---

## Scoring Sheet (for TA/Instructor Use)

| Group Members | C1 (20%) | C2 (20%) | C3 (15%) | C4 (15%) | C5 (15%) | C6 (15%) | **Total** | **Grade** |
|:--------------|:--------:|:--------:|:--------:|:--------:|:--------:|:--------:|:---------:|:---------:|
|               |          |          |          |          |          |          |           |           |
|               |          |          |          |          |          |          |           |           |
|               |          |          |          |          |          |          |           |           |

**Notes:**