---
title: "Regression‑Discontinuity Designs"
subtitle: "PS 312"
author: "Jaye Seawright"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(knitr)
library(rdrobust)    # Robust inference for RD designs
```

## Today's Roadmap

1. **Hook & Activation:** The logic of regression‑discontinuity  
2. **Concept Introduction:** Cutoffs, running variables, and local randomness  
3. **Code Walk-Through** – Does winning a close election pull representatives toward the center or the extremes?  
4. **Diagnostics & Assumptions:** Manipulation checks and placebo tests  
5. **Core Graded Activity:** Write your paragraph for the TA  
6. **Wrap‑Up:** When to use RDD and its modern extensions  

**Goal:** Move from "I've heard of RDD" to "I can design, run, and critically evaluate a regression‑discontinuity analysis in R."

---

class: inverse, center, middle

# 1. Hook & Activation  
### The Logic of Regression‑Discontinuity

---

## Scenario: Merit Scholarships and College Completion

Imagine a state program that awards a $5,000 college scholarship to any high school student who scores at least 1200 on the SAT.

- You could compare college completion rates for scholarship recipients vs. non‑recipients.  
- What's the problem with that? (Hint: Students with higher SAT scores are fundamentally different.)  
- You could compare students who scored *just above* 1200 to students who scored *just below* 1200.  
- Are these two groups of students likely to be similar in other ways?  

The **regression‑discontinuity design (RDD)** exploits the fact that, near the cutoff, whether a student receives the scholarship is as good as random. This allows us to estimate the causal effect of the scholarship by comparing outcomes just on either side of the cutoff.

---

## The RDD Intuition

1.  **Running variable (forcing variable):** A continuous variable that determines treatment assignment (e.g., SAT score).  
2.  **Cutoff:** A threshold value of the running variable. Units above the cutoff receive the treatment; units below do not (or vice versa).  
3.  **Local randomization:** For units *very close* to the cutoff, whether they fall just above or just below is essentially random. These units should be comparable on all other characteristics.  
4.  **Estimation:** Compare the average outcome for units just above the cutoff to the average outcome for units just below the cutoff. The difference is the local average treatment effect (LATE).

---

class: inverse, center, middle

# 2. Concept Introduction  
### Cutoffs, Running Variables, and Local Randomness

---

## The Key Assumption: No Manipulation of the Running Variable

> Units cannot precisely control their value of the running variable to sort themselves onto one side of the cutoff.

- In the scholarship example: Can students precisely manipulate their SAT score to land at 1201 instead of 1199?  
- In an election context: Can candidates precisely manipulate the vote margin to win by 0.1% instead of losing by 0.1%?  

If manipulation is possible, the "as‑if random" comparison breaks down. We can test for manipulation by examining the density of the running variable around the cutoff.

---

## The RDD Model

A common way to estimate an RDD is with a local linear regression:

`$$Y_i = \beta_0 + \beta_1 \text{Treat}_i + \beta_2 (X_i - c) + \beta_3 \text{Treat}_i \times (X_i - c) + \epsilon_i$$`

Where:  
- `\(Y_i\)` is the outcome for unit *i*.  
- `\(X_i\)` is the running variable.  
- `\(c\)` is the cutoff value.  
- `\(\text{Treat}_i = 1\)` if `\(X_i \ge c\)`, and 0 otherwise.  
- `\(\beta_1\)` is the **RDD estimate**—the jump in the outcome at the cutoff.

We can also use more flexible local polynomial methods (via `rdrobust`) that automatically select the optimal bandwidth and provide robust inference.

---

class: inverse, center, middle

# 3. Code Walk-Through  
### Does Winning a Close Election Pull Representatives Toward the Center or the Extremes?

---

Similar to last time's code review, I'm going to show you the code I've used to solve an RDD problem. You as a class are my graders. Give me a grade on the A-F scale, using the rubric on the next slide.

---

| **Criterion** | **Excellent (A)** | **Good (B)** | **Fair (C)** | **Poor (D/F)** |
| :------------ | :---------------- | :----------- | :----------- | :------------- |
| **Correctness** | Code runs without errors; specification correct. | Runs but has minor warnings; RDD logic mostly sound. | Runs with significant warnings or fails; key RDD elements missing. | Does not run or completely incorrect. |
| **Clarity & Documentation** | Clear comments; meaningful variable names; easy to follow. | Some comments; variable names okay. | Few comments; cryptic names. | No comments; incomprehensible. |
| **RDD Best Practices** | Checks manipulation; uses robust inference; sensitivity analysis. | Uses `rdrobust` but misses one diagnostic. | Uses simple OLS without robustness. | No understanding of RDD assumptions. |
| **Efficiency** | No unnecessary steps; appropriate use of `tidyverse`. | Slight redundancy. | Noticeable inefficiency. | Extremely inefficient. |

---

## The Research Question

A long‑standing debate in political science asks whether elected officials moderate their positions to appeal to the broader electorate or whether they cater to their party's base. If winning a close election causes a representative to become more ideologically extreme, we would expect to see a "jump" in their DW‑NOMINATE score at the 50% vote threshold.

**Your task:** Use a regression‑discontinuity design to estimate the effect of *barely winning* a U.S. House election (as a Democrat vs. a Republican) on the ideological extremism of the winning candidate. In the example code below, we will look at raw ideology as the outcome variable; you will probably want a different outcome for your final application later on.

---

## Getting Set Up in R

We'll use the `rdrobust` package for estimation and inference, which includes a built‑in dataset of close U.S. House elections merged with DW‑NOMINATE scores.

```{r, eval=FALSE}
# Install required packages if you don't have them
# install.packages("rdrobust")
# install.packages("tidyverse")
```

```{r}
# Load packages
library(rdrobust)
library(tidyverse)
```

---

## Loading and Preparing the Data

We will use MIT Election Lab data contains U.S. House election returns from 1948–2024, along with the winning candidate's DW‑NOMINATE scores.

```{r load_data}
# ----------------------------------------------------------------------
# Step 1: Install and load required packages
# ----------------------------------------------------------------------
# install.packages(c("tidyverse", "rdrobust", "readr"))

library(tidyverse)
library(rdrobust)

# ----------------------------------------------------------------------
# Step 2: Load U.S. House election results (MIT ElEction Lab data 1976-2024)
# ----------------------------------------------------------------------
# Data source: MIT Election Data and Science Lab
elections_raw <- read_csv("data/house1976-2024.csv")

# ----------------------------------------------------------------------
# Step 3: Download DW-NOMINATE scores (VoteView)
# ----------------------------------------------------------------------
url_voteview <- "https://voteview.com/static/data/out/members/HSall_members.csv"
nominate_raw <- read_csv(url_voteview)
```

---

```{r clean_and_merge}
# ----------------------------------------------------------------------
# Step 4: Clean and merge the datasets
# ----------------------------------------------------------------------
# Clean election data: keep only two-party contests (Democrat vs Republican)
elections_clean <- elections_raw %>%
  filter(party %in% c("DEMOCRAT", "REPUBLICAN")) %>%
  group_by(year, state, district) %>%
  # Keep only districts with exactly one Democrat and one Republican
  filter(n_distinct(party) == 2) %>%
  # Calculate vote share and margin
  mutate(
    total_votes = sum(candidatevotes),
    vote_share = candidatevotes / total_votes,
    # Running variable: Democratic margin (positive = Dem win, negative = Rep win)
    margin_dem = case_when(
      party == "DEMOCRAT" & vote_share > 0.5 ~ (vote_share - 0.5) * 100,
      party == "DEMOCRAT" & vote_share < 0.5 ~ -(0.5 - vote_share) * 100,
      party == "REPUBLICAN" & vote_share > 0.5 ~ -(vote_share - 0.5) * 100,
      party == "REPUBLICAN" & vote_share < 0.5 ~ (0.5 - vote_share) * 100
    ),
    winner = candidatevotes == max(candidatevotes)
  ) %>%
  ungroup() %>%
  # Keep only the winning candidate
  filter(winner == TRUE) %>%
  select(year, state, district, party, margin_dem, candidate_name = candidate)

```

---

```{r clean_and_merge_2}
# Clean DW-NOMINATE data: get first dimension score for each member
nominate_clean <- nominate_raw %>%
  filter(chamber == "House") %>%
  select(congress, state_abbrev, district_code, party_code, 
         bioname, nominate_dim1) %>%
  # Approximate election year from congress number (congress * 2 + 1786)
  mutate(year = 1786 + congress * 2)

# Merge: match by state, district, and approximate election year
# (We'll use fuzzy matching on year because election year and congress year differ slightly)
close_elections <- elections_clean %>%
  mutate(
    state_abbrev = state.abb[match(str_to_title(state), state.name)],
    district_num = as.numeric(district)
  ) %>%
  left_join(
    nominate_clean,
    by = c("state_abbrev", "district_num" = "district_code"),
    suffix = c("_elec", "_nom")
  ) %>%
  filter(abs(year_elec - year_nom) <= 2) %>%  # Match congress within 2 years of election
  # Keep only close elections (within 10 percentage points)
  filter(abs(margin_dem) <= 10) %>%
  mutate(
    treat_dem = if_else(margin_dem >= 0, 1, 0),
    ideology = nominate_dim1
  ) %>%
  distinct(year_elec, state_abbrev, district_num, .keep_all = TRUE)  # Remove duplicate matches
```

---

```{r run_main_rdd, out.width="40%"}
# ----------------------------------------------------------------------
# Step 5: Run the RDD analysis
# ----------------------------------------------------------------------
# Visualize with rdplot
rdplot(
  y = close_elections$ideology,
  x = close_elections$margin_dem,
  c = 0,
  p = 1,
  x.label = "Democratic Vote Margin (%)",
  y.label = "DW-NOMINATE Score",
  title = "Ideology by Election Outcome"
)

# Estimate with rdrobust
rdd_results <- rdrobust(
  y = close_elections$ideology,
  x = close_elections$margin_dem,
  c = 0,
  p = 1,
  kernel = "triangular"
)

summary(rdd_results)
```

---

## Checkpoint: Does Our Design Make Sense?

Having run the RDD, consider:

- **Running variable:** `margin_dem` – the Democratic candidate's margin of victory. A value of 0 is the cutoff.  
- **Cutoff:** 0. Districts where `margin_dem > 0` elected a Democrat; districts where `margin_dem < 0` elected a Republican.  
- **Outcome:** `ideology` – the ideological position of the winning candidate, regardless of party.  
- **Key assumption:** Candidates cannot precisely manipulate the vote margin to fall just above or just below 0. We'll test this.


---

## Interpreting the Results

The `rdrobust` output provides several key quantities:

- **Coefficient:** The estimated jump in `ideology` at the cutoff. A positive value would (counterintuitively) mean that *barely winning* a seat as a Democrat (instead of a Republican) leads to a *more conservative* representative.  
- **Standard error and p‑value:** For hypothesis testing.  
- **Bandwidth (`h`):** The window of data used for estimation (selected to minimize MSE).  
- **Effective number of observations:** How many elections fell within the chosen bandwidth.

**Questions to answer:**  
- What is the sign and magnitude of the effect?  
- Is the effect statistically significant?  
- How would you interpret this substantively? For example: "Winning a close House election as a Democrat increases the winner's DW‑NOMINATE score by X points compared to winning as a Republican."

---

class: inverse, center, middle

# 4. Diagnostics & Assumptions  
### Manipulation Checks and Placebo Tests

---

## Testing for Manipulation: Density of the Running Variable

The most important assumption of RDD is that units cannot precisely manipulate the running variable. In close elections, if candidates could perfectly sort themselves into winning, we would see a "bunching" of observations just above the cutoff and a "gap" just below.

We can test this visually by inspecting a histogram.

```{r running_var_hist}
# Histogram of the running variable
runninghist <- ggplot(close_elections, aes(x = margin_dem)) +
  geom_histogram(binwidth = 0.5, boundary = 0, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Density of Democratic Vote Margin",
    x = "Democratic Vote Margin (%)",
    y = "Number of Elections"
  ) +
  theme_minimal()
```

---

```{r show_running_var_hist, echo=FALSE}
runninghist
```

---

**What to look for:**  
- The histogram should be smooth around the cutoff. No sudden jumps or dips.  
- If there is a discontinuity in the density, manipulation may be a concern.

---

## Placebo Test: Covariate Balance

If the RDD is valid, predetermined covariates should be balanced around the cutoff. We can test this by running the RDD on a covariate that *should not* be affected by the treatment—for example, the district's number.

```{r placebo_test}
# Placebo test on district population
rdd_placebo <- rdrobust(
  y = close_elections$district_num,      # Placebo outcome
  x = close_elections$margin_dem,
  c = 0
)
```

---

```{r show_placebo_test}
summary(rdd_placebo)
```

---

If the placebo test yields a statistically significant effect, it suggests that the discontinuity in our main outcome may be driven by pre‑existing differences rather than the treatment.

---

## Alternative Bandwidth Choices

`rdrobust` selects the bandwidth that minimizes the mean squared error. You can examine how sensitive your results are to different bandwidth choices.

```{r bandwith_sensitivity}
# Estimate with a half and double bandwidth
rdd_half <- rdrobust(
  y = close_elections$ideology,
  x = close_elections$margin_dem,
  c = 0,
  h = rdd_results$bws[1,1] / 2
)

rdd_double <- rdrobust(
  y = close_elections$ideology,
  x = close_elections$margin_dem,
  c = 0,
  h = rdd_results$bws[1,1] * 2
)
```

---

```{r bandwith_sensitivity_2}
# Compare coefficients
coef_half <- rdd_half$coef[1]
coef_optimal <- rdd_results$coef[1]
coef_double <- rdd_double$coef[1]

c(Half = coef_half, Optimal = coef_optimal, Double = coef_double)
```

If the estimate changes dramatically with the bandwidth, the result may be fragile. A stable estimate across reasonable bandwidths is reassuring.

---

So, what's my grade?

---

class: inverse, center, middle

# 5. Core Graded Activity  
### Write Your Paragraph for the TA

---

## Instructions

**By the end of class today, email your TA a short paragraph that includes:**

1. Your **research question** (one sentence).  
2. A brief description of the **RDD design** (running variable, cutoff, treatment, outcome).  
3. The **key result** (the RDD coefficient from `rdrobust`) and its **interpretation**.  
4. An assessment of the **key assumption** (no manipulation of the running variable). Did the density plot look smooth around the cutoff?  
5. **One robustness check** you performed (e.g., placebo test, alternative bandwidth) and what it suggested.

---

## Example Paragraph (for a different question)

> *Our group asks: Does winning a state lottery affect an individual's likelihood of filing for bankruptcy? We use a sharp regression‑discontinuity design with lottery ticket numbers as the running variable and a cutoff at the winning number. Our `rdrobust` estimate indicates that winning the lottery reduces the probability of bankruptcy within five years by 2.3 percentage points (p = 0.01). A density test of ticket numbers shows no evidence of manipulation around the winning threshold. As a placebo test, we checked whether lottery winners differed from losers in pre‑existing credit scores; the RDD estimate was near zero and insignificant, supporting the validity of the design. The results are robust to halving or doubling the bandwidth.*

---

## Reminders

- One submission per student.

---

class: inverse, center, middle

# 6. Wrap‑Up  

> **Single most important rule for RDD:** Always check for manipulation of the running variable. Plot the density and look for smoothness around the cutoff. If the density jumps, the design is suspect.
