---
title: "Regression Kickoff: Social Media & Violence on Jan 6"
subtitle: "PS 312"
author: "Jaye Seawright"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(broom)
library(scales)
library(ggplot2)
```

## Today's Activity

- **Goal:** Use regression to understand how electoral support for Trump in 2016 relates to domestic terrorism in subsequent years.
- **Deliverable:** One table, one figure, and a short paragraph per group.
- **Why regression?** It estimates the **strength** and **uncertainty** of relationships while adjusting for confounders.

To warm up, we'll work through a regression example with the **January 6 defendants data**—the same data we used to build DAGs last time.

---

## The Data: January 6 Defendants

```{r load-data, echo=TRUE}
# Read the data (adjust path if needed)
df_raw <- read_csv("data/final_merged_data.csv")

# Quick look at the variables we care about
df_raw %>%
  select("Case ID", Age, contains("SocialMedia"), contains("ChargedWithViolence")) %>%
  head()
```

---

**Challenge:** Three coders (`.jaye`, `.evelyn`, `.mia`) each recorded whether a defendant used social media and whether they were charged with violence. We need a single, combined measure. And we have a messy group membership variable.

---

## Combining Multiple Coders

Use `coalesce()` to take the first non‑missing value across the three coder columns. And clean up the group membership variable with 'case_when()'.

```{r combine-coders}
df_clean <- df_raw %>%
  mutate(
    SocialMedia = coalesce(SocialMedia.jaye, SocialMedia.evelyn, SocialMedia.mia),
    ChargedWithViolence = coalesce(ChargedWithViolence.jaye, ChargedWithViolence.evelyn, ChargedWithViolence.mia),
    Age = as.numeric(if_else(Age == "#", NA_character_, Age)),
    
    # Create binary group membership
    GroupMember = case_when(
      str_detect(`Group affiliation`, regex("no\\s*known|unknown", ignore_case = TRUE)) ~ 0,
      is.na(`Group affiliation`) | `Group affiliation` == "" ~ 0,
      TRUE ~ 1
    )
  ) %>%
  filter(!is.na(SocialMedia), !is.na(ChargedWithViolence))
```

---

```{r combine-coders-2}
# Check the combined variables
df_clean %>%
  count(SocialMedia, ChargedWithViolence) %>%
  spread(ChargedWithViolence, n, fill = 0)

# Check the group count
df_clean %>% count(GroupMember)
```

Now we have a clean dataset ready for regression.

---


## Simple Regression: Social Media → Violence

Our research question: **Does social media use increase the likelihood of engaging in violence during the January 6 attack?**

Because `ChargedWithViolence` is binary (0/1), we use logistic regression.

```{r simple-logit}
mod1 <- glm(ChargedWithViolence ~ SocialMedia,
            data = df_clean,
            family = binomial)

tidy(mod1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

- **Interpretation:** The odds ratio tells us how much the odds of a violence charge increase for social media users.
- But this relationship may be **confounded** by age (younger defendants use social media more and may be more impulsive).

---

## Multiple Regression: Add a Confounder (Age)

Recall our DAG from last time: **Age** affects both social media use and violence.

```{r multiple-logit}
mod2 <- glm(ChargedWithViolence ~ SocialMedia + Age,
            data = df_clean,
            family = binomial)

tidy(mod2, conf.int = TRUE, exponentiate = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

**Observations:**
- The coefficient on `SocialMedia` may change after controlling for `Age`.
- The adjusted odds ratio gives us an estimate **closer to the causal effect** (under the DAG's assumptions).

---
## Multiple Regression: Add Two Confounders

Now we control for both **Age** and **Group Membership**, following our DAG.

```{r multiple-logit-2}
mod3 <- glm(ChargedWithViolence ~ SocialMedia + Age + GroupMember,
            data = df_clean,
            family = binomial)

tidy(mod3, conf.int = TRUE, exponentiate = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```


---

## Visualizing Results

A coefficient plot makes comparisons easy.

```{r coef-plot, fig.width=7, fig.height=4}
plotme <- mod3 %>%
  tidy(conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  geom_pointrange(size = 1) +
  scale_x_log10(labels = comma) +
  labs(
    title = "Odds Ratios for Violence Charge",
    x = "Odds Ratio (log scale)",
    y = NULL,
    caption = "Dashed line at OR = 1 (no effect)"
  ) +
  theme_minimal(base_size = 14)
```

---

```{r coef-plot-2, fig.width=7, fig.height=4}
plotme
```

---

## Predicted Probabilities

What does the model imply for a 25‑year‑old vs. a 50‑year‑old?

```{r predicted-prob, echo=TRUE}
new_data <- crossing(
  SocialMedia = c(0, 1),
  GroupMember = c(0, 1),
  Age = c(25, 50)
)

probsoutcome <- new_data %>%
  mutate(
    pred_prob = predict(mod3, newdata = ., type = "response")
  ) %>%
  arrange(Age, GroupMember, SocialMedia)
```

---

```{r predicted-prob-2, echo=TRUE}
probsoutcome
```

- **Takeaway:** Social media use might be associated with a lower predicted probability of a violence charge, and age maybe lowers the chance of violence, too. Group membership substantially increases the chance of violence.

---

## Your Turn

1. Load the Trump vote & terrorism data (resources on the class page).
2. Build a regression model that answers the question:
   *How does 2016 Trump support relate to domestic terrorism incidents?*
3. Consider **confounders** (e.g., population, region, economic conditions).
4. Produce:
   - One **regression table** (coefficients with standard errors / p‑values)
   - One **figure** (coefficient plot or marginal effects plot)
5. Write a short paragraph interpreting the key result.
6. Email to your TA.

**Resources:**  
[Class regression help page](https://jnseawright.github.io/PS312/InClass/Regression.html)

```{r, eval=FALSE}
# Example code for your analysis
model <- lm(terrorism_incidents ~ trump_vote_share + confounder1 + confounder2,
            data = your_data)
summary(model)
```
```