---
title: "Difference-In-Differences In-Class Activity PS 312"
author: "Jaye Seawright"
date: "April 23, 2026"
output: html_document
---


Your task for today is to use a difference-in-differences model to test whether transition to democracy increases the risk of interstate war. 

Start by loading the quality of governance data into R Studio:

```{r read data etc}
#Add libraries to load and process data
#install.packages("devtools")
#devtools::install_github("ropengov/rqog")
library(rqog)
#Read in the Quality of Governance dataset
qog_ts <- read_qog(which_data = "standard", data_type = "time-series")
```

Use the codebook on the QOG website to find a variable that measures interstate military conflict and a variable that measures a country's level of democracy. Make sure that your variable measuring democracy is coded as 0 for authoritarian country-years and 1 for democratic country-years. This may require some modification.

Carry out a difference-in-differences analysis using code modeled after the example below. (I am using different treatment and outcome variables, obviously.)
 

```{r example difindif}
qog_ts$central_bank_independence <- 1*(qog_ts$cbi_cbiu>0.5)

library(fect)

example_difindif <- fect(wdi_unempilo ~ central_bank_independence, index = c("cname", "year"), 
                         force = "two-way", data = qog_ts, method="ife",
                         min.T0=1, se=1, na.rm = TRUE)

example_difindif
```


Your job is to assemble an analysis that makes sense to you as a group, interpret the results, and decide together what it does and does not tell us about the causal relationship between democracy and interstate conflict. Send your results and a paragraph you write about your conclusions to your TA!



